Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-08-01 paint-contour-tree.tex
updating with proof of buffered shrubs, and cleaning up reweighting argument

@@

@@

\begin{corollary}\label{cor:geometric}
 Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$.
 Then for any $v \in d$, $|H_v| = O(|H_p|)$.
\end{corollary}

@@

\newcommand{\buff}{\mathcal{B}}
\newcommand{\loss}{\lambda}
\subsection{Pruning away buffered sub-shrubs}\label{sec:buffSub}
Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.

Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths). The remaining portion of the shrub will then be handled in a later subsection. We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.

@@

\begin{definition}\label{def:buff}
 Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
  \begin{enumerate}
  \item $\sum_{p\in \buff} |\loss(p)| \geq \frac{1}{100} \sum_{p\in \buff} |p|$
  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
 \end{enumerate}
\end{definition}

Ultimately we will prune buffered subtrees from $\cU$ by starting at the leaves and working towards the root, and hence the second condition of buffered will trivially be satisfied.

The following will allow buffered sub-shrubs to be charged to their tall support chains.

\begin{lemma}
 Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
\end{lemma}
\begin{proof}
\textcolor{red}{\huge Proof missing. Potentially useful observations below}

 Let $r$ be the root of $\buff$. First observe that $\loss(r) \geq |r|/100$.  Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.  However, this would contradict the second condition of \Def{buff}.

 Consider any node $q\in \buff$.  Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.

 \begin{align*}
  \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
  &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
  &= ????   &\mbox{buffered is not sufficient to get to next line} \\
  &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
  &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
 \end{align*}

 \Cor{geometric} should be used somewhere.
\end{proof}

@@

\newcommand{\pruned}{\cU/\cB}
\subsection{Paying for Pruned Shrubs: the root is everything} \label{sec:weight}

\begin{definition}\label{def:pruningAShrub}
 Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
 We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first. We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
\end{definition}

The goal of this section is to prove the following lemma, which ultimately allows us to pay for pruned shrubs.

@@

\textcolor{red}{\huge The rest of this subsection needs updating}

We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
Note that $R_p$ can be negative. 
Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.

\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
\end{claim}

\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
Traverse $\pmax(T)$ from $q'_i$ to $q$.
Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
do not affect its reduced heap.
The vertices of $\redH_{q'_i}$ that are deleted
are exactly those present in the path $q$. 
Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
theses reduced heaps do not have an ancestor-descendant relationship.
Therefore when $R_q$ is negative, it is at most by $2|q|$.
We sum over all $q$ to complete the proof.
\end{proof}

\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.

 and traverse $\pmax(T)$ up to $p$. Suppose we encounter
the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
are removed from $H_p$ are exactly the vertices that lie in these paths. 
We split this contribution into tall and short paths, observing that the latter
are all unsupported. Hence, we can (crudely) upper bound the number of vertices
removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
}

The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
We stress that the all the previous work is mostly the setup for this claim.

\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
\end{claim}

\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.

After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
\end{proof}

\begin{corollary} \label{cor:chargeCor}
 Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
 Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
\end{corollary}
\begin{proof}
 Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
 By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
\end{proof}

@@

\begin{proof}[Proof of \Lem{cu-root}]
We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
\[
% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
 \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
 \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
\]

We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
Suppose not. Let $p, p'$ be the counterexample furthest from the root,
where $p$ is the parent of $p'$. 
Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
That completes the whole proof.
\end{proof}

@@

\begin{proof} 
For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
removed in the pruning process of \Def{pruningAShrub} that produced the buffer-pruned sub-shrub $\pruned$
(i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
Thus we can decompose the sum as follows, 
\begin{align*}
 \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| 
 &\leq \sum_{\cU\in \cF} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|
 = \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  \sum_{\cU\in \cF}\sum_{\buff\in B(\cU)} \sum_{p \in \buff} \sum_{v \in p} \log |H_v|\\
 &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| \right)\\
 &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \pmax} |p|\log |p|\right),
\end{align*}
where the last two lines follow from \Cor{disjointBuff} and \Lem{pathBounds}, respectively.

>>>>

\newcommand{\loss}{\lambda}

@@

Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.

@@

\begin{corollary}\label{cor:geometric}
 Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$.
 Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
\end{corollary}

\begin{proof} By the previous lemma, $|H_w| = O(|H_p|)$. Note that there are at most
$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $v$
(which is unsupported). Thus, for any $v \in \lambda(p)$, $H_v \geq 99|H_p|/100$,
completing the proof.
\end{proof}

@@

\newcommand{\buff}{\mathcal{B}}
\subsection{Pruning away buffered sub-shrubs}\label{sec:buffSub}
Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths). The remaining portion of the shrub will then be handled in a later subsection. We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.

@@

\begin{definition}\label{def:buff}
 Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
  \begin{enumerate}
  \item $\sum_{p\in \buff} |\loss(p)| > \frac{1}{100} \sum_{p\in \buff} |p|$
  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
 \end{enumerate}
\end{definition}

The following is quite straightforward. By pruning from the leaves towards
the root, we can remove all buffered sub-shrubs from a given $\cU$.

\begin{lemma} \label{lem:split-buff} Any shrub $\cU$ can be partitioned
into $\{\cU', \cB_1, \cB_2, \ldots \cB_r\}$ where each $\cB_i$ is buffered.
Furthermore, no subtree of $\cU'$ is buffered.
\end{lemma}

\begin{proof} Take any reverse topological order of the nodes in $\cU$
(so descendants occur before ancestors). Initialize $\cU'$ to $\cU$.
We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$. (Note that $\cU'$ is redefined
to be the remainder.)

At the end of this process, we have the desired decomposition. Since we process
in reverse topological order, the minimality condition of \Def{buff}
automatically holds. Thus, each pruned sub-shrub is buffered and the remaining $\cU'$
has the desired property.
\end{proof}

We prove an important lemma. The cost of heap operations corresponding 
to a buffered sub-shrub to be charged to tall support chains.

\begin{lemma} \label{lem:buff}
 Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
\end{lemma}
\begin{proof} The proof is done by a charging argument. For every $p \in \buff$
and $v \in \loss(p)$, we create 100 tokens. Each of these tokens is labeled $\log |H_v|$,
and this is called the ``value" of the token. Observe that the sum of token values is, up to a constant
factor, the bound we wish to show. The set of tokens created for all $v \in \loss(p)$ is denoted
$T_p$.

There is a grand total of $100 \sum_{p \in \buff} |\loss(p)|$ tokens, which
(by the buffered property of $\buff$) is strictly more than $\sum_{p \in \buff} |p|$.
We will place all these tokens on the nodes in $\buff$, using the following process.
We reserve topologically sort (descendants before ancestors) the nodes in $\buff$.
For each $p$ in order, we will place tokens in $T_p$ on the nodes of the subtree $\buff_p$.
Take an arbitrary token in $T_p$, and find the smallest (in the ordering) descendant $q$
of $p$ that currently has strictly less than $|q|$ tokens. Place the token on $q$,
and repeat. This is repeated until all descendants $q$ have $|q|$ tokens, or
all tokens in $T_p$ are used up. When a node $q$ has $|q|$ tokens, we use
``$q$ is full".

Now for a useful claim. \emph{For non-root $p$, all tokens of $T_p$ are used up.}
Suppose not. Consider the first such $p$ in the ordering, and look at the situation
when we finish processing $T_p$. Every descendant $q$ of $p$ must be full. Furthermore,
all the tokens on $q \in \buff_p$ must have come from some $T_{p'}$, where $p' \in \buff_p$.
Thus, $\sum_{q \in \buff_p} |q| < |\bigcup_{p' \in \buff_p} T_{p'}|$ $=100\sum_{p' \in \buff_p} |\loss(p')|$. This violates the second condition of buffered shrubs.

Let $r$ be the root of $\buff$. We argue that $T_r$ is \emph{not used up}.
Suppose the contrary, so $T_r$ is used up. By the previous paragraph,
all tokens are used up. But this implies that $\sum_{p \in \buff} |p| \geq |\bigcup_{p \in \buff} T_p|$
$ = 100\sum_{p \in \buff} |\loss(p)|$, violating the buffered condition.

Since $T_r$ is not used up, it must mean that all $p$s are full. Consider any $p$,
and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
for some $v \in \loss(\hat{p})$, for some ancestor $\hat{p}$ of $p$. By \Cor{geometric},
for every $w \in p$, $\log |H_w| = O(\log |H_v|)$. Because $p$ is full,
there are exactly $|p|$ tokens on $p$. Thus, $\sum_{w \in p} \log |H_w|$
is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. We sum
over all $p$ to complete the proof.

% \textcolor{red}{\huge Proof missing. Potentially useful observations below}
%  
%  Let $r$ be the root of $\buff$. First observe that $\loss(r) \geq |r|/100$.  Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.  However, this would contradict the second condition of \Def{buff}.
%  
%  Consider any node $q\in \buff$.  Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
%  
%  \begin{align*}
%   \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
%   &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
%   &= ????   &\mbox{buffered is not sufficient to get to next line} \\
%   &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
%   &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
%  \end{align*}
%  
%  \Cor{geometric} should be used somewhere.
\end{proof}

@@

\newcommand{\pruned}{\cU'}
\subsection{Paying for Pruned Shrubs: the root is everything} \label{sec:weight}

We focus attention on the remaining shrub left by \Lem{split-buff}: this is 
a shrub $\cU$ such that no subtree of $\cU$ is buffered. Refer to such
a shrub as \emph{buffer-pruned}.

% 
% \begin{definition}\label{def:pruningAShrub}
%  Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
%  We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first. We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
% \end{definition}
% 
The goal of this section is to prove the following lemma.
% , which ultimately allows us to pay for pruned shrubs.

@@

We need the following technical claim. Let $N(\pruned)$ denote the set of nodes in $\pruned$.

\begin{claim} \label{clm:weights} There exists a function $\alpha:N(\pruned) \mapsto \RR^+$
with the following properties. (We use $\alpha_p$ to denote the function value.)
Let $A_p = \sum_{q \in \pruned_p} \alpha_q$.
Then, $A_p - 3\sum_{q \in \pruned_p, q \neq p} |q| \leq |H_p| \leq A_p$.
\end{claim}

\begin{proof} We define $\alpha$ inductively, from leaves upwards.
For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider internal node $p$
with children $q_1, q_2, \ldots, q_k$. The heap $H(p)$ is formed by the union
of the $H(q_i)$ heaps, after deleting certain vertices from them, and potentially
adding some new vertices. The new vertices are potentially those that come
from other nodes that have been pruned off.
The vertices in each $q_i$ as well as $\lambda(q_i)$ are deleted.
Define $\alpha_p$ to be the number of new vertices added to $H_p$.

Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
$- \textrm{\# vertices deleted from $H_q$s}$.
The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
There is a factor $2$ because each vertex appears in at most $2$ heaps.

It is evident that $|H_p| \leq \sum_{q \in \pruned_p, q \neq p} \alpha_q = A_p$.
Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} (|q| + |\lambda(q)|)$.
Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(p)|
\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
\end{proof}

We have the main charging claim. This assumes that $A_p$ is increasing exponentially
as we go up the tree. That property is proven later by induction, using
the following claim for sub-shrubs.

\begin{claim} \label{clm:charge} Fix any path $p \in \pruned\setminus r$. Suppose for any $q,q' \in \pruned_p$
where $q$ is a parent of $q'$ in $\pruned_p$, $A_q \geq (5/4)A_{q'}$. Then $\sum_{q \in \pruned_p} |q| \leq A_p/20$.
\end{claim}

\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|A_q|}/99$.
We prove that $\sum_{q \in \pruned_p} \sqrt{|A_q|}/99 \leq A_p/20$
by a charge redistribution scheme. Assume that each $q \in \pruned_p$
starts with $\sqrt{A_q}/99$ units of charge. We redistribute this charge over all
nodes in $\pruned_q$, and then calculate the total charge. For $q \in \pruned_p$, spread
its charge to all nodes in $\pruned_q$ proportional to $\alpha$ values. In other words,
give $(\sqrt{A_q}/99)\cdot(\alpha_{q'}/A_q)$ units of charge to each $q' \in \pruned_q$.

After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\pruned_p$,
denoted $a_0, a_1, a_2, \ldots, a_k$, contributes to the charge at $q$. The charge is expressed in the following
equation. Ae use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 
$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (5/4)^{i/2} \leq \alpha_q/20 $$
The total charge is $\sum_{q \in \pruned_p} \alpha_p/20 = A_p/20$. 
\end{proof}

\begin{claim} \label{clm:exp-A} For any $q, q' \in \pruned_p$ where $q$
is a parent of $q'$, $A_q \geq (4/3)A_{q'}$.
\end{claim}

\begin{proof} We prove this by induction over the depth of $q$.
When $q$ is a leaf, this is vacuously tree. Consider some internal node $q$,
with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
By \Clm{weights}, $|H_{q'_i}| \leq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
to complete the proof.
\end{proof}

The proof of \Lem{cu-root} just combines all these claims, and essentially
repeats the argument in the previous claim. We apply \Clm{exp-A} and \Clm{charge}
to every child of the root $r$. Denoting the children as $q_1, q_2, \ldots$,
we get that $\sum_{q' \in \pruned_{q_i}} |q| \leq A_{q_i}/20$.
Summing this over the children, $\sum_{p \in \pruned, p \neq r} \leq A_r/20$.
By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r$.
We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.
% 
% 
% 
% \begin{proof}[Proof of \Lem{cu-root}] We first pro 
% We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
% then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
% \[
% % \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
%  \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
%  \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
% \]
% 
% We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
% where $p$ is the parent of $p'$. 
% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
% Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
% Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
% That completes the whole proof.
% \end{proof}
% 
% 
% \textcolor{red}{\huge The rest of this subsection needs updating}
% 
% We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
% $q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
% By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
% Note that $R_p$ can be negative. 
% Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
% where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
% Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
% 
% 
% \begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
% \end{claim}
% 
% \begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
% Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
% Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
% Traverse $\pmax(T)$ from $q'_i$ to $q$.
% Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
% do not affect its reduced heap.
% The vertices of $\redH_{q'_i}$ that are deleted
% are exactly those present in the path $q$. 
% Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
% theses reduced heaps do not have an ancestor-descendant relationship.
% Therefore when $R_q$ is negative, it is at most by $2|q|$.
% We sum over all $q$ to complete the proof.
% \end{proof}
% 
% \ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
% $D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
% to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
% 
% 
%  and traverse $\pmax(T)$ up to $p$. Suppose we encounter
% the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
% are removed from $H_p$ are exactly the vertices that lie in these paths. 
% We split this contribution into tall and short paths, observing that the latter
% are all unsupported. Hence, we can (crudely) upper bound the number of vertices
% removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
% Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
% }
% 
% 
% The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
% We stress that the all the previous work is mostly the setup for this claim.
% 
% 
% \begin{corollary} \label{cor:chargeCor}
%  Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
%  Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
% \end{corollary}
% \begin{proof}
%  Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
%  By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
% \end{proof}
% 
%

@@

@@

\begin{proof} Apply \Lem{split-buff} to $\cU$ to get the family of buffered
sub-shrubs $B(\cU)$ and buffer-pruned $\pruned$.
% For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
% remo that produced the buffer-pruned sub-shrub $\pruned$
% (i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
% Thus we can decompose the sum as follows, 
\begin{align*}
 \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| 
 &\leq \sum_{\cU\in \cF} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|
 = \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  \sum_{\cU\in \cF}\sum_{\buff\in B(\cU)} \sum_{p \in \buff} \sum_{v \in p} \log |H_v|\\
 &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| \right)\\
 &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \pmax} |p|\log |p|\right),
\end{align*}
where the last two lines follow from \Cor{disjointBuff} and \Lem{pathBounds}, respectively.
